---
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    self_contained: no
    fig_caption: yes
    highlight: tango
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../../slides.css
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
title: PBM data analysis
subtitle: IBIS challenge 2024
font-family: Garamond
transition: linear
editor_options: 
  chunk_output_type: console
---

```{r setup, eval=TRUE, echo=FALSE, include=FALSE}
library(knitr)
library(kableExtra)
library(rprojroot)
# library(formattable)

root_dir = rprojroot::find_rstudio_root_file()
knitr::opts_knit$set(root.dir = root_dir)
#setwd(root_dir)

# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
	eval = TRUE,
	echo = TRUE,
	fig.align = "center",
	fig.height = 5,
	fig.path = "figures/PBM_",
	fig.width = 7,
	message = FALSE,
	warning = FALSE,
	comment = "",
	results = TRUE,
	root.dir = root_dir,
	size = "tiny"
)
options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

options(width = 300)
# options(encoding = 'UTF-8')

# getwd()
```

```{r param}
#### Define parameters ####
data_type <- "PBM"
board <- "leaderboard"

```

```{r load_metadata}
## Load metadata
metadata_file <- file.path("metadata", board, paste0("TF_DATASET_", data_type, ".tsv"))
setwd(root_dir)
# getwd()
metadata <- read.table(metadata_file, header = TRUE, sep="\t", comment.char = "")
colnames(metadata)[1] <- "TF"
rownames(metadata) <- metadata$DATASET
# View(metadata)
# colnames(metadata)

## Separate the two normalization methods for each sample
norm_methods <- c("SD", "QNZS")
metadata[, c("norm_method", "sample")] <- t(as.data.frame(strsplit(x = metadata$DATASET, split = "_")))

TFS <- unique(metadata$TF)
datasets <- unique(metadata$dataset)
norm_methods <- unique(metadata$norm_method)
samples <- unique(metadata$sample)
```

```{r load_data}

## Initialise with the first sample
#sample <- samples[1]
setwd(root_dir)
#### Load all the data #### 
message("Loading PBM data")
pbm_data <- list()
for (i in 1:nrow(metadata)) {
  #### Load one dataset ####
  TF <- metadata[i, "TF"]
  dataset  <- metadata[i, "DATASET"]
  sample <- metadata[i, "sample"]
  message("\tLoading data\tTF: ", TF, "; dataset: ", dataset, "; sample: ", sample)
  data_file <- file.path("data", board, 'train', "PBM", TF, paste0(dataset, ".tsv"))
  pbm_data[[dataset]] <- read.table(data_file, header = TRUE, sep = "\t", comment.char = "")
  names(pbm_data[[dataset]])[1] <- sub(pattern = "X.", replacement = "", x = names(pbm_data[[dataset]])[1])
  

  ## Sort dataset by spot ID
  spot_IDs <- pbm_data[[dataset]]$id_spot
  pbm_data[[dataset]] <- pbm_data[[dataset]][order(spot_IDs),]
  row.names(pbm_data[[dataset]]) <- pbm_data[[dataset]]$id_spot
  # dim(pbm_data[[dataset]])

}

# View(pbm_data)
# View(pbm_data[[1]])

```

## Signal intensities

### Impact of normalization method (SD vs QNZS)

```{r SD_vs_QNZS, fig.width=10, fig.height=10, out.width="100%", fig.cap="Comparison of mean signal intensities between normalisation methods SD and QNZS. "}

#### Analyse distribution of signal ####

sample <- samples[1]

par(mfrow=c(2,2))

for (sample in samples) {
  
  TF <- unique(metadata[metadata$sample == sample, "TF"])
  
  SD <- pbm_data[[paste0("SD_", sample)]]
  QNZS <- pbm_data[[paste0("QNZS_", sample)]]
  spot_IDs <- sort(unique(c(SD$id_spot, QNZS$id_spot)))
  
  ## Make sure the two datasets have the same spot IDS
  SD <- SD[spot_IDs, ]
  QNZS <- QNZS[spot_IDs, ]
  
  ## Check that the PBM sequence is the same between SD and QNZS datasets
  table(SD$pbm_sequence == QNZS$pbm_sequence)
  
  ## Compare signal between normalisation methods
  plot(SD$mean_signal_intensity, QNZS$mean_signal_intensity,
       main = paste(TF, sample),
       xlab = "SD signal (log scale)",
       ylab = "QNZS signal",
       log="x",
       col="#0044FF",
       panel.first = grid(col= "#888888"))
}
par(mfrow=c(1,1))


```

## Distribution of signal intenties

### Normalization of SD data

We normalize the SD data by computing a z-score, with outlier-robust estimators of central tencendy (median) and dispersion (IQR).

$$z = \frac{x -\text{median}(X)}{\text{IQR(X)}/1.34898}$$ The scaling factor ($1.34898$) normalises the IQR to obtain an estimator of the theoretical standard deviation of the Gaussian distribution.

```{r SD_normalisation, fig.width=15, fig.height=10, out.width="100%", fig.cap="**Distribution of z-scores from SD-normalised data.** After log10-transformation of the mean signal intensities, the z-score is computed from robust estimators of central tendency (median) and dispersion (IQR). The dark green curve fits the Normal distribution. Vertical bars indicate the mean (plain) and standard deviations (dotted) of the z-score distribution. Note that the shape of the histogram shows strong variations depending on the dataset.  "}
norm_method <- "SD"

#for (method in norm_methods)  {

par(mfrow=c(2,2))

for (sample in samples) {
  dataset <- paste(sep="_", norm_method, sample) 
  TF <- unique(metadata[metadata$dataset == dataset, "TF"])

  attach(pbm_data[[dataset]])
  log10_signal <- log10(mean_signal_intensity)
  
  ## Use the median as estimator of central tendency, because it is robust to outliers
  m_est <- median(log10_signal)

  ## Use standardized IQR as estimator of dispersion (sd), because it is robust to outliers
  iqr <- IQR(log10_signal)
  s_est <- iqr / (qnorm(p = 0.75) - qnorm(p = 0.25))
  z <- (log10_signal - m_est) / s_est
  # sd(z)
  # mean(z)
  # median(z)
  # IQR(z)

  ## Plot an histogram of the z-score values
  h <- hist(z, 
            breaks=100, 
            ylab="Number of PBM spots",
            col="#BBDDFF",
            main=paste(TF, dataset),
            las=1)
  
  
  ## Fit a normal curve 
  h_area <- sum((h$breaks[2:length(h$breaks)] - h$breaks[1:(length(h$breaks)-1)]) * h$counts)
  x = seq(-4, +4, 0.1)
  lines(x, dnorm(x) * h_area, lwd=2, col="darkgreen")
  abline(v = 0, col="darkgreen", lwd=1.5)
  abline(v = c(-1, 1), col="darkgreen", lwd=1.5, lty="dotted")

  message(paste(collapse="\t", norm_method, TF, dataset, 4, sum(z > 4)))
  detach(pbm_data[[dataset]])
  pbm_data[[dataset]]$log10_signal <- log10_signal
  pbm_data[[dataset]]$z <- z
}

par(mfrow=c(1,1))

```

### Distributions of SD signal intensities before and after normalisation

```{r signal_distrib_SD, fig.width=15, fig.height=10, out.width="100%", fig.cap="Distribution of mean signal intensities. "}
#### Plot the distribution of mean signal intensities ####


sample <- samples[1]

par(mfrow=c(4,3))
method <- "SD"

#for (method in norm_methods)  {

for (sample in samples) {
  
  dataset <- paste(sep = "_", method, sample)
  TF <- unique(metadata[metadata$dataset == dataset, "TF"])
  
  
  ## Histogram of the SD-normalized signal intensities
  signal <- pbm_data[[dataset]]$mean_signal_intensity
  # plot(sort(signal, decreasing = TRUE))
  quantiles <- quantile(x = signal, probs = seq(0,1, 0.05))
  q95 <- quantile(x = signal, probs = 0.95, na.rm = TRUE)
  hist(signal, breaks=1000, 
       # xlim=c(0, q95),
       col="#BBBBBB", border="#666666",
       main=paste(TF, sample, "\nSD signal"))
  
  ## histogram of log10-transformed signal intensities
  signal_log10 <- log(signal, base=10)
  # plot(sort(signal_log10, decreasing = TRUE))
  # quantiles_log10 <- quantile(x = signal_log10, probs = seq(0,1, 0.05))
  breaks_log10 <- seq(1, 4.5, 0.05)
  hist(signal_log10,
       breaks=breaks_log10, 
       ylab="Number of PBM spots",
       col="#DDBBFF", border="#8866BB",
       main=paste(TF, dataset),
       las=1)
  
  
  ## histogram of the z-score values
  z <- pbm_data[[dataset]]$z
  h <- hist(z, 
            breaks=100, 
            ylab="Number of PBM spots",
            col="#BBDDFF", border="#6688BB", 
            main=paste(TF, dataset),
            las=1)
  
  
  ## Fit a normal curve 
  h_area <- sum((h$breaks[2:length(h$breaks)] - h$breaks[1:(length(h$breaks)-1)]) * h$counts)
  x = seq(-4, +4, 0.1)
  lines(x, dnorm(x) * h_area, lwd=2, col="darkgreen")
  abline(v = 0, col="darkgreen", lwd=1.5)
  abline(v = c(-1, 1), col="darkgreen", lwd=1.5, lty="dotted")

  # hist(signal_log10, 
  #      breaks=breaks_log10, 
  #      main=paste(TF, method, sample, "\nlog10(signal)"))
  # bg_log10 <- log(pbm_data[[dataset]]$mean_background_intensity)
  # #plot(signal_log10, bg_log10)
  # hist(signal_log10 - bg_log10, 
  #      breaks=100, 
  #      main=paste(TF, method, sample, "\nlog10(signal/bg)"))

    message(paste(collapse="\t", TF, dataset, 1000, sum(signal >1000)))
}

  
  
par(mfrow=c(1,1))


```

### QNZS-normalized data

```{r signal_distrib_QNZS, fig.width=15, fig.height=10, out.width="100%", fig.cap="Distribution of mean signal intensities. "}
#### Plot the distribution of mean signal intensities ####


sample <- samples[1]

par(mfrow=c(2,2))
method <- "QNZS"

#for (method in norm_methods)  {

for (sample in samples) {
  
  dataset <- paste(sep = "_", method, sample)
  TF <- unique(metadata[metadata$dataset == dataset, "TF"])
  
  signal <- pbm_data[[dataset]]$mean_signal_intensity
  # summary(signal)
  # plot(sort(signal, decreasing = TRUE))
  quantiles <- quantile(x = signal, probs = seq(0,1, 0.05))
  q95 <- quantile(x = signal, probs = 0.95, na.rm = TRUE)
  
  ## Compare signal between normalisation methods
  h <- hist(signal, breaks=1000, 
       col='#BBFFDD', border = '#66BB88', 
       las=1, 
       main=paste(TF, sample, "\nQNZS signal"))
  abline(v=seq(-5, 10, 1), col="gray", lwd=1)
  abline(h=seq(0, 250, 50), col="gray", lwd=1)
  abline(v=4, col="red", lwd=2)
#  abline(v=2, col="dark", lwd=2)
  
  ## Fit a normal curve on the histogram
  h_area <- sum((h$breaks[2:length(h$breaks)] - h$breaks[1:(length(h$breaks)-1)]) * h$counts)
  x = seq(-4, +4, 0.1)
  lines(x, dnorm(x) * h_area, lwd=2, col="brown")
  abline(v = 0, col="brown", lwd=1.5)
  abline(v = c(-1, 1), col="brown", lwd=1.5, lty="dotted")

  ## Fit a normal curve with robust estimators on the histogram
  h_area <- sum((h$breaks[2:length(h$breaks)] - h$breaks[1:(length(h$breaks)-1)]) * h$counts)
  x <- seq(-4, +4, 0.1)
  s <- IQR(signal) / (qnorm(p=0.75) - qnorm(p=0.25))
  m <- median(signal)
  lines(x, dnorm(x, 
              mean = m, 
              sd = s) * h_area, 
        lwd=2, col="darkgreen")

  message(paste(collapse="\t", TF, dataset, 4, sum(signal > 4)))

}

  
  
par(mfrow=c(1,1))


```
