---
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    self_contained: no
    fig_caption: yes
    highlight: tango
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../../slides.css
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
title: Analysis of sequence scanning results
subtitle: IBIS challenge 2024
font-family: Garamond
transition: linear
editor_options: 
  chunk_output_type: console
---

```{r setup, eval=TRUE, echo=FALSE, include=FALSE}
library(knitr)
library(kableExtra)
library(rprojroot)
library(data.table)
library(pROC)
# library(formattable)

root_dir = rprojroot::find_rstudio_root_file()
knitr::opts_knit$set(root.dir = root_dir)
#setwd(root_dir)

# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
	eval = TRUE,
	echo = TRUE,
	fig.align = "center",
	fig.height = 5,
	fig.path = "figures/classif-stats",
	fig.width = 7,
	message = FALSE,
	warning = FALSE,
	comment = "",
	results = TRUE,
	root.dir = root_dir,
	size = "tiny"
)
options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

options(width = 300)
# options(encoding = 'UTF-8')

# getwd()
```

```{r param}
#### Define default parameters ####
board <- 'leaderboard'
experiment <- "CHS"
TF <- "GABPA"
dataset <- "THC_0866" 
```

## Metadata

```{r load_metadata}
## Load metadata
metadata_file <- file.path("metadata", board, "TF_DATASET_all-types.tsv")
setwd(root_dir)
# getwd()
metadata <- read.table(metadata_file, header = TRUE, sep="\t", comment.char = "")
colnames(metadata)[1] <- "TF"
rownames(metadata) <- metadata$DATASET
# View(metadata)

TFS <- unique(metadata$TF)
datasets <- unique(metadata$DATASET)
experiments <- unique(metadata$EXPERIMENT)

```

```{r metadata_summary}
metadata_summary <- table(metadata[, c("TF", "EXPERIMENT")])
metadata_summary <- cbind(metadata_summary, total=apply(metadata_summary, 1, sum))
metadata_summary <- metadata_summary[order(metadata_summary[, "total"], decreasing = TRUE), ]
kable(metadata_summary, caption="Summary of the metadata. Number of datasets per transcription factor (rows) and experiment (colums)")
```


## Datasets

- **PCM:** we scanned sequences with position-count matrices (PCM) resulting from motif discovery in in train sequence sets. Motif discovery was performed with  `peak-motifs`  using 3 algorithms: `oligo-analysis`, `dyad-analysis` and `position-analysis`. 

- **sequences:** we scan three types of sequence sets. 

    - **train:** training sequences. Note that for PBM the training sequences are the 500 spots having the highest signal intensity, whereas for other experiments (CHS, GHTS, HTS, SMS) the train sequences are all the sequences downloaded from the IBIS web site ([leaderboard](https://ibis.autosome.org/download_data/leaderboard) or [final](https://ibis.autosome.org/download_data/final)).  

    - **rand:** random genome fragments of the same sizes as the train sequences (see previous section)

    - **test:** test sequences downloaded from the IBIS web site ([leaderboard](https://ibis.autosome.org/download_data/leaderboard) or [final](https://ibis.autosome.org/download_data/final)).  


```{r load_data}
#### Define a function to load a data table and analyse it ####

#' @title Load matrix-scan results for one dataset of the IBIS challenge. 
#' @description Load matrix-scan results for one dataset of the IBIS challenge. These results are tab-delimited files for different sequence types (by default: train, rand and test)
#' @author Jacques van Helden
#' @param dataset the ID of the dataset (must be present in the metadata table)
#' @param seq_types=c("train","rand","test") sequence types for which the hit tables have to be read
#' @examples
#' ## Load recount metadata
#' load_dataset(dataset=THC_0866)
#' @return a list with the data + statistics per motif (motif)
#' @export
load_dataset <- function(dataset,
                         seq_types = c("train", "rand")
#                         seq_types = c("train", "rand", "test")
                         ) {
  ## Initialize variables
  TF <- metadata[dataset, "TF"]
  experiment <- metadata[dataset, "EXPERIMENT"]

  top_hits <- list()
  top_hits$dataset <- dataset
  top_hits$TF <- TF
  top_hits$experiment <- experiment
  top_hits$board <- board

  
  message("Loading sequence scan tables for dataset ", dataset, "\t", TF, "\t", experiment)
  message("Sequence types ", paste(collapse = ",", seq_types))
  
  scan_dir <- file.path("results", 
                        board,
                        "train", 
                        experiment, 
                        TF, 
                        dataset, "peak-motifs-nopurge", "clustered_motifs", "", "matrix-clusters_aligned_logos", "All_concatenated_motifs_trimmed-info_0.1", "sequence-scan")
  
  message("\t", scan_dir)
  
  ## Read the data tables for the different sequence types
  for (seq_type in seq_types) {
    scan_file <- file.path(scan_dir, 
                           paste0(experiment, "_",
                                  TF, "_",
                                  dataset, 
                                  "_peakmo-clust-matrices_", 
                                  seq_type,
                                  ".tsv.gz"))
    
    message("\t", seq_type, "\t", basename(scan_file))
    
    system.time(top_hits[[seq_type]] <- 
                  data.table::fread(
                    cmd = paste0("zless ", scan_file, " | grep -v '^;' | grep -v 10000000"), 
                    sep = "\t", 
                    header = TRUE, 
                    stringsAsFactors = FALSE, 
                    showProgress = TRUE))
    top_hits[[seq_type]]$seq_type <- seq_type
    ## View(top_hits[[seq_type]])
  }

  
  return(top_hits)
}

```

```{r ckassif_statistics}

#' @title Compute classification statistics from two data.frames corresponding to positive and negative cases respectively. 
#' @description Compute classification statistics from two data.frames corresponding to positive and negative cases respectively. The positive and negative data frames must contain a column with numeric scores, which will server to rank them and compute the classification statistics (TP, TN, FP, FN and derived statistics). Another column must be specified to indicate some grouping (in our case, by motif). 
#' 
#' For the IBIS challenge, the positive and negative datasets are feature tables produced by matrix-scan (keeping only the columns 1, 3-6 and 8). The grouping column is the ft_name column, which indicates the motif name. Statistics are computed separately for each motif in the input data table. 
#' @author Jacques van Helden
#' @param pos_data a data frame containing the positive cases
#' @param neg_data a data frame containing the negative cases
#' @param score_column="weight" column of pos_data and neg_data containing the scores §by default, weight score of matrix-scan)
#' @param motif_column=2 column containing the grouping criterion (by default, the ft_name column of matrix-scan)
#' @param decreasing=TRUE if TRUE, score column is sorted by decreasing order, i.e. the highest scores are considered the best. 
#' @return a list with different objects
#' @export

classif_statistics <- function(pos_data, 
                               neg_data,
                               label_column="seq_type",
                               pos_label="train",
                               neg_label="rand",
                               score_column="weight",
                               motif_column = "ft_name",
                               decreasing = TRUE) {

  message("Computing classification statistics")
  message("\tPositive data dimensions: ", nrow(pos_data), " x ", ncol(pos_data))
  message("\tNegative data dimensions: ", nrow(pos_data), " x ", ncol(pos_data))
  merged <- data.frame(rbind(pos_data, 
                  neg_data)) # merge train and rand
  message("\tmerged table before cleaning: ", nrow(merged), " x ", ncol(merged))
  
  ## Sort merged data by score
  merged <- merged[order(merged[, score_column], decreasing = decreasing), ]
  
  ## Suppress very low scores corresponding to sequences with not a single hit
  merged <- merged[merged$weight > -1000, ]
  message("\tmerged table after cleaning: ", nrow(merged), " x ", ncol(merged))
  
  motifs <- unique(merged[, motif_column])
  nmotifs <- length(motifs)
  
  #### Compute statistics for each motif #### 
  cals_stats_per_motif <- function(x) {
    result <- c(
      "min" = min(x),
      "max" = max(x),
      "mean" = mean(x),
      "median" = median(x)
    )
    return(result)
  }

  #### Compute basic descriptive stat about each motif ####
  stats_per_motif <- do.call(data.frame, aggregate(weight ~ ft_name, data = merged, FUN = cals_stats_per_motif))
  #stats_per_motif <- subset(stats_per_motif, select = -1)
  stats_per_motif$pos <- do.call(data.frame, aggregate(weight ~ ft_name, data = pos_data, FUN = cals_stats_per_motif))
  stats_per_motif$neg <- do.call(data.frame, aggregate(weight ~ ft_name, data = neg_data, FUN = cals_stats_per_motif))
  
  stats_per_motif <- do.call(data.frame, stats_per_motif)
  row.names(stats_per_motif) <- stats_per_motif$ft_name
  colnames(stats_per_motif)
  rownames(stats_per_motif)
  stats_per_motif <- 
    subset(x = stats_per_motif, 
           select = grep(pattern = "ft_name", 
                         x = colnames(stats_per_motif),
                         invert = TRUE))
  
  # View(stats_per_motif)
  
  #### Compute distributions of scores per motif ns derive classification statistics ####
  score_table <- list()
  roc_curves <- list()
  
  motif <- motifs[1] ## initialise for testing
  i <- 0
  for (motif in motifs) {
    i <- i + 1
    message("\t\t", dataset, "\t", TF, "\t", experiment, "\t", motif)
    stats <- merged[merged$ft_name == motif, ]
    
    stats$true_label <- stats[,label_column] == pos_label
    stats$pos <- cumsum(stats[,label_column] == pos_label) # number of positive     
    stats$neg <- cumsum(stats[,label_column] == neg_label) # number of positive     
    npos <- sum(stats[,label_column] == pos_label) # number of positive cases
    nneg <- sum(stats$seq_type == neg_label)  # number of negative cases
    # View(stats)
    # head(stats)
    # tail(stats)
    stats$rank <- 1:nrow(stats)
    stats$TP <- cumsum(stats$seq_type == "train")
    stats$FP <- cumsum(stats$seq_type == "rand")
    stats$FN <- npos - stats$TP
    stats$TN <- nneg - stats$FP
    
    ## True Positive Rate (TPR) = Sensitivity = Coverage = Recall
    stats$TPR <- stats$TP / (stats$TP + stats$FN)
    
    ## False Positive Rate (TPR)
    stats$FPR <- stats$FP / (stats$FP + stats$TN)

    ## Positive Predictive Value (PPV) = Precision
    stats$PPV <- stats$TP / (stats$TP + stats$FP)
    
    #### Compute stat per score (aggregate counts)
    if (decreasing) {
      new_score <- c(TRUE, stats[2:nrow(stats), score_column] 
                     < stats[1:(nrow(stats)-1), score_column])
    } else {
      new_score <- c(TRUE, stats[2:nrow(stats),score_column] 
                     > stats[1:(nrow(stats)-1),score_column])
    }
    
    score_table[[motif]] <- stats[new_score, c(6, 8:ncol(stats))]
    
    #### Compute the area under the ROC curve ####
    nscores <- nrow(score_table[[motif]])
    range(score_table[[motif]][2:nscores, "FPR"])
    ## X axis interval
    x <- c(0, score_table[[motif]][, "FPR"])
    y <- c(0, score_table[[motif]][, "TPR"])
    xdiff <- x[2:length(x)] - x[1:(length(x)-1)]
    ## Curve height
    ymean <- (y[2:length(x)] + y[1:(length(x)-1)])/2

    stats_per_motif[motif, "AuROC"] <- as.vector(xdiff %*% ymean)
    
    #### Compute the AuRoC using pROC package ####
    # table(true_labels)
    roc_curves[[motif]] <- pROC::roc(
      as.numeric(stats$true_label), 
      as.vector(stats[, score_column]), 
      direction="<")
    auc_value <- auc(roc_curves[[motif]])
    stats_per_motif[motif, "pROC.auc"] <- auc_value
    if (i==1) {
      plot(roc_curves[[motif]], col = i, lty = i)
    } else {
      lines(roc_curves[[motif]], col = i, lty = i)
    }
  }  
  
  
  result <- list(
    experiment = experiment,
    TF = TF,
    dataset = dataset,
    top_hits = top_hits,
    motifs = motifs,
    stats_per_motif = stats_per_motif,
    score_table = score_table
  )
  return(result)  
}
  
```


```{r weight_hist, fig.width=8, fig.height=8, out.width="100%", fig.cap="Weight distributions for a given dataset"}
plot_weights <- function(results,
                         pos_label="train",
                         neg_label="rand") {
  Wpos <- top_hits[[pos_label]]$weight
  Wneg <- top_hits[[neg_label]]$weight
  
  ## suppress weights corresponding to no hits (N sequences)
  Wpos <- Wpos[Wpos > -1000]
  Wneg <- Wneg[Wneg > -1000]
  
  Wmin <- floor(min(c(Wpos, Wneg))/10)*10
  Wmax <- ceiling(max(c(Wpos, Wneg))/10)*10
  
#  par(mfrow=c(2,1))    
  hist(Wpos, 
       breaks = seq(from=Wmin, to=Wmax, by=0.5),
       main=paste0(top_hits$experiment, " ", 
                   top_hits$TF, " ",
                   pos_label, "\n", dataset),
       xlab = "Weight score (top hit per sequence)",
       ylab = "Number of sequences",
       las=1,
       col = "palegreen")
       
  hist(Wneg, 
       breaks = seq(from=Wmin, to=Wmax, by=0.5),
       main=paste0(top_hits$experiment, " ", 
                   top_hits$TF, " ",
                   pos_label, "\n", dataset),
       xlab = "Weight score (top hit per sequence)",
       ylab = "Number of sequences",
       las=1,
       col = "pink")
       
#  par(mfrow=c(1,1))    

}
```


```{r ROC, fig.width=8, fig.height=8, out.width="100%", fig.cap="ROC for a given dataset"}

plot_curves <- function(classif_stat,
                        type="ROC") {
#  attach(classif_stat)
  ## Initialise parameters for this result set
  dataset <- classif_stat$dataset
  TF <- classif_stat$TF
  experiment <- classif_stat$experiment
  stats_per_motif <- classif_stat$stats_per_motif
  motifs <- classif_stat$motifs
  nmotifs <- length(motifs)
  
  if (type == "ROC") {
    xstat <- "FPR"
    ystat <- "TPR"
    legend_pos <- "bottomright"
#    performance_stat <- "AuROC"
    performance_stat <- "pROC.auc"
    
  } else if (type == "Precision/recall") {
    xstat <- "PPV"
    ystat <- "TPR"
    legend_pos <- "topleft"
#    performance_stat <- "AuROC"
    performance_stat <- "pROC.auc"
  } else {
    stop('Invalid curve type for plot_curves(). Supported types :  "ROC",  "Precision/recall"')
  }
  
  ## Extract performance per motif
  # perf <- classif_stat$stats_per_motif[, performance_stat]
  stats_per_motif$rank <- nmotifs + 1 - rank(classif_stat$stats_per_motif[, performance_stat])
  
  # stats_per_motif[, c("AuROC", "rank")]

  ## Sort motifs by increasing performance to draw the most performing 
  ## one at the end, so it appears above the other ones
  motif_order <- order(stats_per_motif[, performance_stat], decreasing = FALSE)
  sorted_motifs <- rownames(stats_per_motif)[motif_order]

  stats_per_motif$color <- stats_per_motif$rank
  stats_per_motif$lwd <- 2 - 1.5*(stats_per_motif$rank / nmotifs)
  stats_per_motif$lty <- stats_per_motif$rank

  message("Drawing ", type, " curve\t", experiment, "\t", TF, "\t", dataset, "\t", nmotifs, " motifs")
  motif <- sorted_motifs[1]
  stats <- classif_stat$score_table[[motif]]
  
  plot(stats[, xstat], 
       stats[, ystat], 
       lwd = stats_per_motif[motif, "lwd"],,
       col = stats_per_motif[motif, "color"],
       lty = stats_per_motif[motif, "lty"],
       
       main = paste0(type, ": ",
                     experiment, " ", TF, "\n", dataset, " (", nmotifs, " motifs)"),
       
       xlab = xstat, 
       xlim = c(0,1),
       ylab = ystat,
       ylim = c(0,1),

       type = "l",
       las = 1,
       panel.first = c(
         abline(h=seq(0, 1, 0.1), col="#DDDDDD"),
         abline(v=seq(0, 1, 0.1), col="#DDDDDD"),
         abline(v = c(0, 1), col = "black"),
         abline(h = c(0, 1), col = "black")
       )
  )
  
  if (type == "ROC") {
    abline(a = 0, b = 1, col = "black")
  }
  
  ## Draw lines
  for (i in 2:nmotifs) {
    motif <- sorted_motifs[[i]]
    stats <- classif_stat$score_table[[motif]]
    lines(stats[, xstat],
          stats[, ystat], 
          lwd = stats_per_motif[motif, "lwd"],,
          col = stats_per_motif[motif, "color"],
          lty = stats_per_motif[motif, "lty"])
  }
  
  
  perf <- round(digits = 3, 
                stats_per_motif[sorted_motifs, performance_stat])
  legend(legend_pos,
         legend = rev(paste0(sorted_motifs, " (", perf, ")")), 
         col = rev(stats_per_motif[sorted_motifs, "color"]), 
         lwd = rev(stats_per_motif[sorted_motifs, "lwd"]),
         lty = rev(stats_per_motif[sorted_motifs, "lty"]),
         cex = min(0.8, 10/nmotifs))
#  detach(classif_stat)

}
  
```


```{r analysis, fig.width=8, fig.height=8, out.width="100%"}
# View(metadata)
dataset <- "YWL_B_AffSeq_A11_ZNF407.C4" ## For testing
dataset <- "THC_0364.Rep-DIANA_0293"
dataset <- "UT380-127-3"
for (dataset in datasets) {
  TF <- metadata[dataset, "TF"]
  experiment <- metadata[dataset, "EXPERIMENT"]

  if (experiment == "PBM") {
    message ("Skipping PBM dataset ", dataset)
    next()
  }

  message("Analysing dataset ", dataset, "\t", TF, "\t", experiment)
  
  ## Load the positive and negative datasets
  top_hits <- load_dataset(dataset)
  
  ## Compute performance statistics
  classif_stat <- classif_statistics(
    pos_data = top_hits[["train"]],
    neg_data = top_hits[["rand"]])

  
  ## Check that the manual computation of AuROC corresponds to the result of pROC::auc
  plot(classif_stat$stats_per_motif[, c("AuROC",
       "pROC.auc")], 
       xlim=c(0,1),
       ylim = c(0,1), 
       col = "blue",
       panel.first = abline(a=0,b=1))
  
  # plot_weights(top_hits)
  #  par(mfrow=c(1,2))
  plot_curves(classif_stat, type = "Precision/recall")
  plot_curves(classif_stat, type = "ROC")
#  par(mfrow=c(1,1))
}

```

